Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
Final version thesis

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence in iterative imputation algorithms. Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\\widehat{R}$ and AC diagnostics. Of course, it never hurts to iterate longer, but such calculations hardly bring added value."
keywords: missing data, iterative imputation, non-convergence, MICE
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
---

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values. The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data? 
<!-- $\dot y_{\rm imp, \ell}$ $y_{\rm mis}$ -->

These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R]. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R`. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences. We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

@@

To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations). The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in $\dot y_{\rm imp, \ell}$.

In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence impacts the distribution of the imputed values in this example, $\dot y_{\rm imp, \ell}$. This effect may translate into the distribution of the completed data {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$ and the pooled estimate $\bar{Q}$. $\bar{Q}$ is then a  biased, invalid estimate of $Q$, since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18].

@@

A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature. This $\theta$ is computed as the estimate in each imputation $\ell$, $\hat{Q}_\ell$, across iterations. Implementing this, however, might be somewhat too technical for empirical researchers. Besides, this scalar summary is not model-independent. That is, this $\theta$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. On these grounds, such a $\theta$ can be considered insufficient too.

@@

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation $\ell$. $\lambda_{1,\ell}$ then summarizes the largest possible amount of covariance in the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. By definition, each $\lambda_{1, \ell}$-value is equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state-space of the algorithm.

@@

In this study, we consider each of the four $\theta$s that were discussed above. Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92] to evaluate mixing, and autocorrelation [$AC$; @scha97; @gelm13] to diagnose trending. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$. Other methods are outside the scope of this study because they e.g. assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.

@@

In 2019, Vehtari et al. proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992. The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity. This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values. Namely, rank-normalization, folding, and localization. The definition of the diagnostic itself is equal to the original $\widehat{R}$. We follow Vehtari et al.'s formulation (2019, p. 5). Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest. For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92]. The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$. Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other. When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied. The variance between chains $B$ is then equivalent to the variance within chains $W$, and $\widehat{R}$-values will be close to one. High $\widehat{R}$-values thus indicate non-convergence.

@@

Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}. Panel A consists of the two traceplots that may be evaluated through visual inspection. Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$. Panel C displays the traditional computation of $\widehat{R}$ conform @gelm92 (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).

@@

When we look at panel B, we conclude something weird. The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$ for both scenarios. However, if we look at the chain means of the non-convergence scenario, there should be signs of trending  up-to iteration number seven. After consulting the documentation on `stats::acf()`, we conclude that this $AC$ function is not suited for iterative imputation algorithms. The function is optimized for performance when $t\geq50$ [@box15], while the default number of iterations in iterative imputation is often much lower. Therefore, we compute $AC$ manually, see the solid line. These $AC$-values behave as expected from visual inspection. We will, therefore, only consider manually calculated $AC$-values as non-convergence diagnostic.

In panel C, we do not see a lot of difference between the two scenarios. The $\widehat{R}$-values are even somewhat lower for the non-convergence scenario than for the typical convergence. And apparently, the large variance causes an increase in $\widehat{R}$-values. Convergence seems to worsen with every additional iteration. This paints the wrong picture. An explanation for this behavior may be that under severe trending, the assumption of over-dispersion is violated. Taken together, this version of $\widehat{R}$ is sub-optimal for assessing non-convergence in MI.

The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up. An exception to this general trend is the 'dip' around $t=3$. As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario. We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four. Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values. For low values of $t$, the $\widehat{R}$-values conform Vehtari et al. are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms. We will compute $\widehat{R}$ conform @veht19, and calculate $AC$ manually. The diagnostics will be applied to the four types of scalar summaries $\theta$ described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix. We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.

@@

The simulation study aims to determine to what extend non-convergence affects iterative imputation algorithms. Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm. And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods. We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$. In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain. Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values). Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.

@@

The missingness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. Higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. Still, we expect that missingness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.

The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence. Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the initial values of the algorithm [which are sampled randomly from the set of observed datapoints; @buur18]. As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state). We expect that we can induce non-convergence by early stopping in at least those conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$ [@mice].

@@

We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 3.6.3; @R]. The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).

@@

Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$. And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$). Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$. In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$. That is, the results are more or less stable for conditions where $T \geq 30$. Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).

@@

3. Compute non-convergence diagnostics $\widehat{R}$ and $AC$. Do *not* use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R]. Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/)).

>>>>

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands.
corrauth: "Hanne Oberman, Sjoerd Groenman building, Padualaan 14, 3584 CH, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence in iterative imputation algorithms. Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\\widehat{R}$ and $AC$ diagnostics. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value."
keywords: missing data, iterative imputation, non-convergence, MICE
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
---

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values. The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data? 
<!-- $\dot y_{\rm imp, \ell}$ $y_{\rm mis}$ -->

These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R]. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package in `R` [@mice]. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences. We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$). The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

@@

To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays two scenarios from the example. The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations). The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\rm imp, \ell}$.

In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\rm imp, \ell}$. This effect may translate into the distribution of the $m$ sets of completed data {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$. $\bar{Q}$ is then a  biased, invalid estimate of $Q$. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.

<!-- , since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. -->

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].

@@

A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature. This $\theta$ is computed as the estimate in each imputation, $\hat{Q}_\ell$, across iterations. Implementing this, however, might be somewhat too technical for empirical researchers. Besides, this scalar summary is not model-independent. That is, $\theta=\hat{Q}$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. On these grounds, such a $\theta$ can be considered insufficient too.

@@

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation. The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.

@@

In this study, we consider each of the four $\theta$s that we discussed. Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$. Other methods are outside the scope of this study because they e.g. assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.

@@

In 2019, Vehtari et al. proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992. The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity. This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values. Namely, rank-normalization, folding, and localization. The definition of the diagnostic itself is equal to the original $\widehat{R}$. Therefore, we may follow Vehtari et al.'s formulation (2019, p. 5) to define both $\widehat{R}$ versions. Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest. For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92]. The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$. Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other. When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied. The variance between chains, $B$, is then equivalent to the variance within chains, $W$, and $\widehat{R}$-values will be close to one. High $\widehat{R}$-values thus indicate non-convergence.

@@

Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}. Panel A consists of the two traceplots that may be evaluated through visual inspection. Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$. Panel C displays the traditional computation of $\widehat{R}$ conform Gelman and Rubin (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).

@@

When we look at panel B, we conclude something weird. The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$ for both scenarios. However, if we look at the chain means of the non-convergence scenario, there should be signs of trending  up-to iteration number seven. After consulting the documentation on `stats::acf()`, we conclude that this $AC$ function is not suited for iterative imputation algorithms. The function is optimized for performance when $t\geq50$ [@box15], while the default number of iterations in iterative imputation is often much lower. Therefore, we compute $AC$ manually (presented with solid lines in panel B). These $AC$-values behave as expected from visual inspection. We will, therefore, only consider manually calculated $AC$-values as non-convergence diagnostic.

In panel C, we do not see a lot of difference between the two scenarios either. The $\widehat{R}$-values are even somewhat lower for the non-convergence scenario than for the typical convergence. And apparently, the large variance causes an increase in $\widehat{R}$-values. Convergence seems to worsen with every additional iteration. This paints the wrong picture. An explanation for this behavior may be that under severe trending, the assumption of over-dispersion is violated. Taken together, this version of $\widehat{R}$ is sub-optimal for assessing non-convergence in MI.

The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up. An exception to this general trend is the 'dip' around $t=3$. As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario. We can explain this as a downside of the low number of iterations in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four. Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values. For low values of $t$, the $\widehat{R}$-values conform Vehtari et al. are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms. We will compute $\widehat{R}$ conform @veht19, and calculate $AC$ manually. The diagnostics will be applied to the four types of scalar summaries $\theta$ described earlier: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix. We evaluate the performance of these eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.

@@

The simulation study aims to determine to what extend non-convergence affects iterative imputation algorithms. Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm. And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods. If the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$, we will conclude that the algorithm is sufficiently converged for practical purposes. In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain. Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values). Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.

@@

The missingness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. A higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. Still, we expect that missingness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.

The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence. Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the starting values [which are sampled randomly from the set of observed datapoints; @buur18]. As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state). We expect that we can induce non-convergence by early stopping in at least those conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$ [@mice].

@@

We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 3.6.3; @R]. The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).

@@

Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$. And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$). Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$. In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$. That is, the results are more or less stable for conditions where $T \geq 30$. Full results are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).

@@

3. Compute non-convergence diagnostics $\widehat{R}$ and $AC$. Do *not* use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R]. Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint)).
